Skew-symmetric entropy stable discontinuous Galerkin formulations

### 2018-08-14 docs/skewform/main.tex
updating skewform paper

\date{}
\author{Jesse Chan}%, David C.\ Del Rey Fernandez}
\title{Skew-symmetric entropy stable discontinuous Galerkin formulations with applications to hybrid and non-conforming meshes}

@@

\begin{abstract}
Entropy stable high order methods for nonlinear conservation laws satisfy an inherent discrete entropy inequality.  The construction of such schemes has relied on the use of carefully chosen collocation points \cite{gassner2013skew, fisher2013high, carpenter2014entropy, chan2018efficient} or volume and surface quadrature rules \cite{chan2017discretely, chan2018discretely} to produce operators which satisfy a summation-by-parts (SBP) property.  In this work, we show how to generalize the construction of semi-discretely entropy stable schemes to accommodate more general pairs of volume and surface quadratures under which an SBP property does not hold.  We conclude by discussing the applicability of these new operators for entropy stable schemes on hybrid and non-conforming meshes.  
\end{abstract}

@@

We begin by reviewing the dissipation of entropy for a $d$-dimensional system of nonlinear conservation laws on a domain $\Omega$
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{j=1}^d\pd{\bm{f}_j(\bm{u})}{x_j} = 0, \qquad \bm{u}\in \mathbb{R}^n, \qquad \bm{f}:\mathbb{R}^n\rightarrow\mathbb{R}^n,
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}$ are the conservative variables and $\bm{f}(\bm{u})$ is a vector-valued nonlinear flux function.  We are interested in nonlinear conservation laws for which a convex entropy function $U(\bm{u})$ exists.  For such systems, the  \emph{entropy variables} are an invertible mapping $\bm{v}(\bm{u}):\mathbb{R}^n\rightarrow \mathbb{R}^n$ defined as the derivative of the entropy function with respect to the conservative variables 
\begin{align}
\bm{v}(\bm{u}) = \pd{U}{\bm{u}}.%, \qquad  \text{ invertible}.%\bm{u}\LRp{\bm{v}(\bm{u})} = \bm{u}.
\label{eq:entropyvarsmap}
\end{align}
Several widely used equations in fluid modeling (Burgers, shallow water, compressible Euler and Navier-Stokes equations) yield convex entropy functions $U(\bm{u})$ \cite{hughes1986new, chen2017entropy}.  Let $\partial \Omega$ be the boundary of $\Omega$ with outward unit normal $\bm{n}$.  By multiplying the equation (\ref{eq:nonlineqs}) with $\bm{v}(\bm{u})^T$, the solutions $\bm{u}$ of (\ref{eq:nonlineqs}) can be shown to satisfy an entropy inequality
\begin{equation}
\int_{\Omega}\pd{U(\bm{u})}{t}\diff{x} + \int_{\partial \Omega} \sum_{j=1}^d \LRp{\bm{v}(\bm{u})^T\bm{f}_j(\bm{u}) - \psi_j\LRp{\bm{v}(\bm{u})}}n_j \diff{x} \leq 0, 
\label{eq:entropyineq}
\end{equation}
where $\bm{n} = \LRp{n_1,\ldots,n_d}$ denotes the outward unit normal, and $\psi_j(\bm{u})$ is some function referred to as the entropy potential.

@@

We also define trace spaces for each reference element.  Let $\hat{f}$ be a face of the reference element $\hat{D}$.  The trace space $V^N \LRp{\hat{f}}$ is defined as the restrictions of functions in $V^N\LRp{\hat{D}}$ to $\hat{f}$, and denote the dimension of the trace space as ${\rm dim}\LRp{V^N\LRp{{\hat{f}}}} = N^f_p$.  
\[
V^N \LRp{\hat{f}} = \LRc{ \left.u\right|_{\hat{f}}, \quad u \in V^N\LRp{\hat{D}}}, \qquad \hat{f}\in \partial\hat{D}.
\]
For example, on a $d$-dimensional simplex, $V^N \LRp{\partial \hat{D}}$ consists of total degree $N$ polynomials on simplices of dimension $(d-1)$.  On a $d$-dimensional tensor product element, $V^N \LRp{\partial \hat{D}}$ consists of maximum degree $N$ polynomials on a tensor product element of dimension $(d-1)$.

@@

\note{TODO: Standardize hat notation for operators/normals: can refer to operators on reference element and do $k$ superscript for physical elements, or do hats on reference element.}
Let $\hat{D} \subset\mathbb{R}^d$ denote a reference element with surface $\partial \hat{D}$.  
The high order schemes in \cite{chan2017discretely, chan2018discretely} begin by approximating the solution in a degree $N$ polynomial basis $\LRc{\phi_j({\bm{x}})}_{i=1}^{N_p}$ on $\hat{D}$.  These schemes also assume volume and surface quadrature rules $({\bm{x}}_i, w_i)$, $\LRp{{\bm{x}}^f_i,w^f_i}$ on $\hat{D}$.  We will specify the accuracy of each quadrature rule later, and discuss how quadrature accuracy implies specific summation-by-parts properties.

@@

The matrix $\bm{Q}^i_N$ also satisfies a decoupled summation-by-parts (SBP) property, which is used to prove semi-discrete entropy stability for nonlinear conservation laws.  
\begin{lemma}
Assume that the volume and surface quadrature rules are sufficiently accurate such that the quantities
\[
\int_{\hat{D}} \pd{u}{x_i} v, \qquad \int_{\partial \hat{D}} u v n_i
\]
are integrated exactly for all $u,v \in V^N\LRp{\hat{D}}$ and $i = 1,\ldots, d$.  Then, the decoupled SBP operator $\bm{D}^i_N$ satisfies a summation by parts property:
\begin{gather}
\bm{Q}^i_N+\LRp{\bm{Q}^i_N}^T = \LRp{\begin{array}{cc}\bm{0}&\\ & \bm{W}_f {\rm diag}(\bm{n}_i)\end{array}} = \bm{B}_N\label{eq:dsbp}
\end{gather}
%\begin{enumerate}
%\item The mass matrix is positive definite, 
%\item The quadrature integrates exactly $\int_{\hat{D}} \pd{u}{x_i} v$ for all $u,v \in V^N\LRp{\hat{D}}$ and $i = 1,\ldots, d$.
%\end{enumerate}
\label{lemma:dsbp}
\end{lemma}
\begin{proof}
The proof is a straightforward extension of Theorem 1 in \cite{chan2017discretely} to polynomial approximation spaces on non-simplicial elements.  
\end{proof}

The assumptions of Lemma~\ref{lemma:dsbp} are satisfied for sufficiently accurate volume and surface quadratures.  For example, on simplicial elements, (\ref{eq:dsbp}) holds if the volume quadrature is exact for polynomial integrands of total degree $(2N-1)$, and the surface integral is exact for degree $2N$ polynomials on each face.  Tensor product elements require stricter conditions: (\ref{eq:dsbp}) holds if the volume quadrature is exact for polynomial integrands of degree $2N$ in each coordinate, due to the fact that derivatives of $u\in Q^N$ are degree $(N-1)$ with respect to one coordinate and degree $N$ with respect to others.

@@

\paragraph{Triangular elements (Figure~\ref{subfig:trigll})} We next consider a triangular element $\hat{D}$, where the volume quadrature is exact for degree $2N$ polynomials \cite{xiao2010quadrature} and an $(N+1)$-point GLL quadrature on each face.  Let $u,v \in P^N$ denote two arbitrary degree $N$ polynomials.  The derivative $\pd{u}{x} \in P^{(N-1})$, and $\pd{u}{x}v \in P^{(2N-1)}$, so the volume quadrature satisfies the assumptions of Lemma~\ref{lemma:dsbp}.  However, because the surface quadrature is exact only degree $(2N-1)$ polynomials and the trace of $uv\in P^{2N}$, the surface quadrature does not satisfy the assumptions of Lemma~\ref{lemma:dsbp}.

@@

These specific pairings of volume and surface quadratures appear naturally for hybrid meshes consisting of DG-SEM quadrilateral elements (using GLL volume quadrature) and triangular elements, as shown in Figure~\ref{fig:hybrid}.  In Figure~\ref{subfig:hybrid1}, the surface quadrature is a $(N+1)$ point GLL rule, and results in a loss of the SBP property on the triangle.  In Figure~\ref{subfig:hybrid2}, the surface quadrature is a $(N+1)$ point Gauss-Legendre rule, and results in a loss of the SBP property on the quadrilateral element.

The goal of this work is to construct high order accurate discretizations which preserve entropy conservation for these situations in which the decoupled SBP property (\ref{eq:dsbp}) does not hold.

@@

The property (\ref{eq:dsbp}) (which we will refer to as the ``strong'' SBP property) relates the polynomial exactness of specific quadrature rules to algebraic properties of quadrature-based matrices.  We will relax accuracy conditions on these quadrature rules by utilizing not the strong SBP property, but a weaker ``variational'' version of the SBP property.

\begin{lemma}
Assume that $v \in V^M$.  %, and let the surface trace $\LRl{v}_{\partial \hat{D}} \in V^M\LRp{\partial \hat{D}}$ for some $M \leq N$. 
If the volume quadrature is exact for polynomials of degree $N+M-1$ and the surface quadrature is exact for polynomials of degree $N+M$, then for an arbitrary vector $\bm{u}_q$, 
\[
\bm{v}_q^T\bm{Q}_i \bm{u}_q = \bm{v}_q^T\LRp{ \LRp{\bm{V}_f\bm{P}_q}^T \bm{W}_f \diag{{\bm{n}}_i}\bm{V}_f\bm{P}_q - \bm{Q}_i^T}\bm{u}_q
\]
where $\bm{v}$ denotes the vector of values of $v(\bm{x})$ of at both volume and surface quadrature points. 
\label{lemma:vsbp}
\end{lemma}
\begin{proof}
Recall that $\bm{Q}_i = \bm{W} \bm{V}_q \bm{D}^i\bm{P}_q$.  Since the volume quadrature is exact for degree $2N-1$ polynomials, we have that
\begin{align*}
\bm{v}_q^T\bm{Q}_i \bm{u}_q &= \bm{v}_q^T\bm{W} \bm{D}^i \bm{P}_q \bm{u}_q = \int_{\hat{D}} \pd{\Pi_N u}{\hat{x}_i} v = \int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i - \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i}.
\end{align*}
Since the volume integrand $\LRp{\Pi_N u} \pd{v}{\hat{x}_i} \in V^{2N-1}$, it is exactly integrated using the volume quadrature.  Additionally, since $\Pi_N u \in V^N$ and $v\in V^M$ on the surface of $\hat{D}$, the surface quadrature is exact for the surface integral and 
\begin{align*}
&\int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i -   \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i} =\\
 &\bm{v}_f^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \LRp{\bm{V}_q\bm{D}^i\bm{P}_q\bm{v}_q}^T\bm{W} \bm{V}_q\bm{P}_q\bm{u}_q\\
=& \bm{v}_q^T \LRp{\bm{V}_f\bm{P}_q}^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \bm{v}_q^T \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q\bm{u}_q,
\end{align*}
where we have used that, since $v\in V^N$, $\bm{v}_f = \bm{V}_f\bm{P}_q\bm{v}_q$.  The proof is completed by noting that 
\[
\bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{M}\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T\bm{W} = \bm{Q}_i^T.
\]
\end{proof}

The variational SBP property also extends to the decoupled SBP operator.  This property, along with the exact differentiation of constants, is necessary for the proof of entropy stability.  
\begin{lemma}
Let $\bm{D}^i_N$ be a decoupled SBP operator on the reference element $\hat{D}$, and let the surface quadrature be exact for polynomials of degree $N+M$ for some $M \leq N$.  Suppose $v\in V^N$ and that $\LRl{v}_{\partial \hat{D}} \in P^M\LRp{\partial \hat{D}}$, Then, 
\[
\bm{v}^T\bm{Q}^i_N\bm{u} = \bm{v}^T\LRp{\bm{B}^i_N - \LRp{\bm{Q}^i_N}^T}\bm{u}.%, \qquad \bm{Q}^i_N \bm{1} = \bm{0}.
\]
where $\bm{v}$ denotes the values of $v$ of at both volume and surface quadrature points.  
\label{lemma:vdsbp}
\end{lemma}
\begin{proof}
For convenience, let $\bm{u}_q, \bm{u}_f$ denote evaluations of $u$ at volume and surface points, such that 
\[
\bm{u} = \begin{pmatrix} \bm{u}_q\\ \bm{u}_f\end{pmatrix}, \qquad \bm{v} = \begin{pmatrix} \bm{v}_q\\ \bm{v}_f\end{pmatrix}.  
\]
The proof of the variational summation by parts property uses the definition of $\bm{Q}^i_N$ (\ref{eq:QN}), 
\begin{align*}
\bm{v}^T\bm{Q}^i_N\bm{u} &= \bm{v}_q^T\bm{Q}_i \bm{u}_q - \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}}_i \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}}_i \bm{u}_f\\
& - \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}}_i \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} + \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}}_i \bm{u}_f.
\end{align*}
Applying Lemma~\ref{lemma:vsbp} then yields
\begin{align*}
\bm{v}^T\bm{Q}^i_N\bm{u} =& -\bm{v}_q^T\bm{Q}^T_i \bm{u}_q + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}}_i \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}}_i \bm{u}_f\\
&\qquad - \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}}_i \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} - \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}}_i \bm{u}_f + \bm{v}_f^T\bm{W}_f\hat{\bm{n}}_i \bm{u}_f\\
=& \begin{pmatrix} \bm{v}_q\\ \bm{v}_f\end{pmatrix}^T 
\left(\begin{pmatrix}
\bm{0}& \\
& \bm{W}_f\diag{\hat{\bm{n}}_i}
\end{pmatrix}\right.\\
&+
\left.\begin{pmatrix}
-\bm{Q}_i^T + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q}^T\bm{W}_f\hat{\bm{n}}_i \bm{V}_f\bm{P}_q & -\frac{1}{2} \LRp{\bm{W}_f\hat{\bm{n}}_i \bm{V}_f\bm{P}_q}^T\\
\frac{1}{2}\bm{W}_f\hat{\bm{n}}_i \bm{V}_f\bm{P}_q & -\frac{1}{2}\bm{W}_f\hat{\bm{n}}_i
\end{pmatrix}  \right)
\begin{pmatrix} \bm{u}_q\\ \bm{u}_f\end{pmatrix}\\
=& \bm{v}^T\LRp{\bm{B}^i_N - \LRp{\bm{Q}^i_N}^T}\bm{u}.
\end{align*}
\end{proof}
Using Lemma~\ref{lemma:vdsbp}, we will determine the accuracy of the surface quadrature necessary to ensure the proof of semi-discrete entropy stability is valid.  This minimal degree of surface quadrature accuracy will depend on the nature of the reference-to-physical mapping (e.g.\ affine vs curved elements).

>>>>

\date{}
\author{Jesse Chan}%, David C.\ Del Rey Fernandez}
\title{Skew-symmetric entropy stable discontinuous Galerkin formulations}

@@

\begin{abstract}
Entropy stable high order methods for nonlinear conservation laws satisfy an inherent discrete entropy inequality.  The construction of such schemes has relied on the use of carefully chosen collocation points \cite{gassner2013skew, fisher2013high, carpenter2014entropy, chan2018efficient} or volume and surface quadrature rules \cite{chan2017discretely, chan2018discretely} to produce operators which satisfy a summation-by-parts (SBP) property.  In this work, we show how to construct skew-symmetric schemes which are entropy stable even for volume and surface quadratures under which an SBP property does not hold.  These skew-symmetric formulations avoid the use of a ``strong'' SBP property, and require only that operators exactly differentiate constants and satisfy a discrete form of the fundamental theorem of calculus.  We conclude by discussing the applicability of these new operators for entropy stable schemes on hybrid meshes.% and non-conforming meshes.  
\end{abstract}

@@

We begin by reviewing the dissipation of entropy for a $d$-dimensional system of nonlinear conservation laws on a domain $\Omega$
\begin{equation}
\pd{\bm{u}_h}{t}  + \sum_{j=1}^d\pd{\bm{f}_j(\bm{u})}{x_j} = 0, \qquad \bm{u}\in \mathbb{R}^n, \qquad \bm{f}:\mathbb{R}^n\rightarrow\mathbb{R}^n,
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}$ are the conservative variables and $\bm{f}(\bm{u})$ is a vector-valued nonlinear flux function.  We are interested in nonlinear conservation laws for which a convex entropy function $U(\bm{u})$ exists.  For such systems, the  \emph{entropy variables} are an invertible mapping $\bm{v}(\bm{u}):\mathbb{R}^n\rightarrow \mathbb{R}^n$ defined as the derivative of the entropy function with respect to the conservative variables 
\begin{align}
\bm{v}(\bm{u}) = \pd{U}{\bm{u}}.%, \qquad  \text{ invertible}.%\bm{u}\LRp{\bm{v}(\bm{u})} = \bm{u}.
\label{eq:entropyvarsmap}
\end{align}
Several widely used equations in fluid modeling (Burgers, shallow water, compressible Euler and Navier-Stokes equations) yield convex entropy functions $U(\bm{u})$ \cite{hughes1986new, chen2017entropy}.  Let $\partial \Omega$ be the boundary of $\Omega$ with outward unit normal $\bm{n}$.  By multiplying the equation (\ref{eq:nonlineqs}) with $\bm{v}(\bm{u})^T$, the solutions $\bm{u}$ of (\ref{eq:nonlineqs}) can be shown to satisfy an entropy inequality
\begin{equation}
\int_{\Omega}\pd{U(\bm{u})}{t}\diff{x} + \int_{\partial \Omega} \sum_{j=1}^d \LRp{\bm{v}(\bm{u})^T\bm{f}_j(\bm{u}) - \psi_j\LRp{\bm{v}(\bm{u})}}n_j \diff{x} \leq 0, 
\label{eq:entropyineq}
\end{equation}
where $\bm{n} = \LRp{n_1,\ldots,n_d}$ denotes the outward unit normal, and $\psi_j(\bm{u})$ is some function referred to as the entropy potential.

@@

The proofs presented in this work will also use anisotropic tensor product polynomial spaces, where the maximum polynomial degree varies depending on the coordinate direction.  We denote such spaces by $Q^{N_1, \ldots, N_d}$, where $N_k$ are non-negative integers and
\[
Q^{N_1, N_2, \ldots, N_d}\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq i_k \leq N_k, \quad k = 1,\ldots, d}.
\]
For example, the isotropic tensor product space $Q^N$ is the same as $Q^{N,\ldots,N}$.

We also define trace spaces for each reference element.  Let $\hat{f}$ be a face of the reference element $\hat{D}$.  The trace space $V^N \LRp{\hat{f}}$ is defined as the restrictions of functions in $V^N\LRp{\hat{D}}$ to $\hat{f}$, and denote the dimension of the trace space as ${\rm dim}\LRp{V^N\LRp{{\hat{f}}}} = N^f_p$.  
\[
V^N \LRp{\hat{f}} = \LRc{ \left.u\right|_{\hat{f}}, \quad u \in V^N\LRp{\hat{D}}, \quad \hat{f}\in \partial\hat{D}}.
\]
For example, on a $d$-dimensional simplex, $V^N \LRp{\partial \hat{D}}$ consists of total degree $N$ polynomials on simplices of dimension $(d-1)$.  On a $d$-dimensional tensor product element, $V^N \LRp{\partial \hat{D}}$ consists of maximum degree $N$ polynomials on a tensor product element of dimension $(d-1)$.

@@

%\note{TODO: Standardize hat notation for operators/normals: can refer to operators on reference element and do $k$ superscript for physical elements, or do hats on reference element.}
Let $\hat{D} \subset\mathbb{R}^d$ denote a reference element with surface $\partial \hat{D}$.  
The high order schemes in \cite{chan2017discretely, chan2018discretely} begin by approximating the solution in a degree $N$ polynomial basis $\LRc{\phi_j({\bm{x}})}_{i=1}^{N_p}$ on $\hat{D}$.  These schemes also assume volume and surface quadrature rules $({\bm{x}}_i, w_i)$, $\LRp{{\bm{x}}^f_i,w^f_i}$ on $\hat{D}$.  We will specify the accuracy of each quadrature rule later, and discuss how quadrature accuracy implies specific summation-by-parts properties.

@@

For sufficiently accurate quadrature rules, the matrix $\bm{Q}^i_N$ also satisfies a decoupled summation-by-parts (SBP) property, which is used to prove semi-discrete entropy stability for nonlinear conservation laws.  
\begin{lemma}
Assume that the volume and surface quadrature rules are sufficiently accurate such that the quantities
\[
\int_{\hat{D}} \pd{u}{x_i} v, \qquad \int_{\partial \hat{D}} u v n_i
\]
are integrated exactly for all $u,v \in V^N\LRp{\hat{D}}$ and $i = 1,\ldots, d$.  Then, the decoupled SBP operator $\bm{D}^i_N$ satisfies a summation by parts property:
\begin{gather}
\bm{Q}^i_N+\LRp{\bm{Q}^i_N}^T = \LRp{\begin{array}{cc}\bm{0}&\\ & \bm{W}_f {\rm diag}(\bm{n}_i)\end{array}} = \bm{B}_N\label{eq:dsbp}
\end{gather}
%\begin{enumerate}
%\item The mass matrix is positive definite, 
%\item The quadrature integrates exactly $\int_{\hat{D}} \pd{u}{x_i} v$ for all $u,v \in V^N\LRp{\hat{D}}$ and $i = 1,\ldots, d$.
%\end{enumerate}
\label{lemma:dsbp}
\end{lemma}
\begin{proof}
The proof is a straightforward extension of Theorem 1 in \cite{chan2017discretely} to polynomial approximation spaces on non-simplicial elements.  
\end{proof}

The assumptions of Lemma~\ref{lemma:dsbp} are satisfied for sufficiently accurate volume and surface quadratures.  For example, on simplicial elements, (\ref{eq:dsbp}) holds if the volume quadrature is exact for polynomial integrands of total degree $(2N-1)$, and the surface integral is exact for degree $2N$ polynomials on each face.  Tensor product elements require stricter conditions: (\ref{eq:dsbp}) holds if both the volume and surface quadratures are exact for polynomial integrands of degree $2N$ in each coordinate, due to the fact that derivatives of $u\in Q^N$ are degree $(N-1)$ with respect to one coordinate and degree $N$ with respect to others.

@@

\paragraph{Triangular elements (Figure~\ref{subfig:trigll})} We next consider a triangular element $\hat{D}$, where the volume quadrature is exact for degree $2N$ polynomials \cite{xiao2010quadrature} and an $(N+1)$-point GLL quadrature on each face.  Let $u,v \in P^N$ denote two arbitrary degree $N$ polynomials.  The derivative $\pd{u}{x} \in P^{(N-1)}$, and $\pd{u}{x}v \in P^{(2N-1)}$, so the volume quadrature satisfies the assumptions of Lemma~\ref{lemma:dsbp}.  However, because the surface quadrature is exact only degree $(2N-1)$ polynomials and the trace of $uv\in P^{2N}$, the surface quadrature does not satisfy the assumptions of Lemma~\ref{lemma:dsbp}.

@@

These specific pairings of volume and surface quadratures appear naturally for hybrid meshes consisting of DG-SEM quadrilateral elements (using GLL volume quadrature) and triangular elements, as shown in Figure~\ref{fig:hybrid}.  In Figure~\ref{subfig:hybrid1}, the surface quadrature is a $(N+1)$ point GLL rule, and results in a loss of the SBP property on the triangle.  In Figure~\ref{subfig:hybrid2}, the surface quadrature is a $(N+1)$ point Gauss-Legendre rule, and results in a loss of the SBP property on the quadrilateral element.  The goal of this work is to construct high order accurate discretizations which preserve entropy conservation for situations in which the decoupled SBP property (\ref{eq:dsbp}) does not hold.

@@

The property (\ref{eq:dsbp}) (which we will refer to as the ``strong'' SBP property) relates the polynomial exactness of specific quadrature rules to algebraic properties of quadrature-based matrices.  We will relax accuracy conditions on these quadrature rules by utilizing not the strong SBP property, but a weaker ``variational'' version of the SBP property.  This variational property relies on the assumption that the volume and surface quadrature rules are exact for polynomials of certain degrees.

\begin{assumption}
\label{ass:quad}
Let $v(\bm{x})$ be a polynomial such that $v \in V^{M_{\rm vol}}$ with surface trace $\LRl{v}_{\partial \hat{D}} \in V^{M_{\rm face}}\LRp{f}$ for $f \in \partial \hat{D}$, where $M_{\rm vol}, M_{\rm face} \leq N$ are two non-negative integers.  We assume that the volume and surface quadrature rules are sufficiently accurate such that
\begin{enumerate}
\item the mass matrix $\bm{M}$ is positive definite, 
\item the volume and surface quadratures exactly integrate the quantities $\int_{\hat{D}} \pd{u}{x_i} v$ and $\int_{\partial \hat{D}} u v n_i$ for all $u \in V^N\LRp{\hat{D}}$, $i = 1,\ldots, d$, and $f \in \partial \hat{D}$.  
\end{enumerate}
\end{assumption}

This section presumes that Assumption~\ref{ass:quad} is satisfied for general  $M_{\rm vol}, M_{\rm surf}$.  In Sections~\ref{sec:affine} and \ref{sec:curved}, specific values of $M_{\rm vol}, M_{\rm surf}$ will be motivated by proofs of entropy stability, and Assumption~\ref{ass:quad} will be verified for specific choices of volume and surface quadrature on simplicial and tensor product elements.
%Suppose Assumption~\ref{ass:quad} is satisfied, and let $\bm{v}_q$ denote the vector of values of $v(\bm{x})$ at volume quadrature points.  Then, for an arbitrary vector $\bm{u}_q$, 
%\[
%\bm{v}_q^T \bm{Q}_i \bm{u}_q = \int_{\hat{D}} \pd{u}{x_i} v, \qquad \LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f \diag{\hat{\bm{n}_i}} \bm{V}_f\bm{P}_q\bm{u}_q =  \int_{\partial \hat{D}} \Pi_N u v n_i,
%\]
%where we have used that, because $v$ is a polynomial of degree $N$, it is equal to its own $L^2$ projection.

The conditions in Assumption~\ref{ass:quad} are sufficient to ensure that the following variational SBP property holds:
\begin{lemma}
\label{lemma:vsbp}
%Assume that $v \in V^M$.  
%If the volume quadrature is exact for polynomials of degree $N+M-1$ and the surface quadrature is exact for polynomials of degree $N+M$, 
Suppose Assumption~\ref{ass:quad} holds, and let $\bm{v}_q$ denote the vector of values of $v(\bm{x})$ of at volume quadrature points. Let $u(\bm{x})$ be an $L^2$ integrable function, and let $\bm{u}_q$ denote the values of $u(\bm{x})$ at quadrature points.  Then, 
\[
\bm{v}_q^T\bm{Q}_i \bm{u}_q = \bm{v}_q^T\LRp{ \LRp{\bm{V}_f\bm{P}_q}^T \bm{W}_f \diag{\hat{\bm{n}}_i}\bm{V}_f\bm{P}_q - \bm{Q}_i^T}\bm{u}_q.
\]
\end{lemma}
\begin{proof}
Recall that $\bm{Q}_i = \bm{W} \bm{V}_q \bm{D}^i\bm{P}_q$.  The quadrature-based $L^2$ projection $\Pi_Nu$ is a polynomial of degree $N$, and is computed discretely as $\bm{P}_q\bm{u}_q$.  By the exactness of the volume quadrature in Assumption~\ref{ass:quad}, 
\begin{align*}
\bm{v}_q^T\bm{Q}_i \bm{u}_q &= \bm{v}_q^T\bm{W} \bm{D}^i \bm{P}_q \bm{u}_q = \int_{\hat{D}} \pd{\Pi_N u}{\hat{x}_i} v = \int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i - \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i}.
\end{align*}
%Since the volume integrand $\LRp{\Pi_N u} \pd{v}{\hat{x}_i} \in V^{N+M-1}$, it is exactly integrated using the volume quadrature.  
%Additionally, since $\Pi_N u \in V^N$ and $v\in V^M$ on the surface of $\hat{D}$, the surface quadrature is exact for the surface integral and 
By the exactness of the surface quadrature in Assumption~\ref{ass:quad}, 
\begin{align*}
&\int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i -   \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i} =\\
 &\bm{v}_f^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \LRp{\bm{V}_q\bm{D}^i\bm{P}_q\bm{v}_q}^T\bm{W} \bm{V}_q\bm{P}_q\bm{u}_q\\
=& \bm{v}_q^T \LRp{\bm{V}_f\bm{P}_q}^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \bm{v}_q^T \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q\bm{u}_q,
\end{align*}
where we have used that, since $v \in V^N$, $\bm{v}_f = \bm{V}_f\bm{P}_q\bm{v}_q$.  The proof is completed by noting that 
\[
\bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{M}\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T\bm{W} = \bm{Q}_i^T.
\]
\end{proof}
